Tail associations in ecological variables and their impact on extinction risk

### 2020-01-19 Response.txt

Dear editors,

Thank you for the attention and expertise you used in assessing, for Ecosphere, our paper "Tail associations and their impact on extinction risk" (this was the old title, we will now be changing it in response to a referee request - see below). Thank you also for the generally very positive reviews. We have responded to all referee comments through changes to the manuscript, and below. We find the paper is improved and we hope you now find it ready for formal acceptance to Ecosphere. Our responses to comments are preceeded by ">>>" to make them easy to find. Thank you again for an enjoyable and helpful review process.

Yours,
Shyamolina Ghosh, Lawrence Sheppard and Dan Reuman

@@

>>>Correct. Thank you for the positive feedback.

I suggest only a few clarifications:

Specific comments
L164-175: Has this particular formulation been analyzed in detail in a classic text? Cite it to guide the reader to more resources on the behavior of this model more generally.

>>>Models of this form (i.e., stochastic matrix models, of which this is an example) are very common in demography, evolutionary ecology, and quantitative conservation biology. As the referee requests, we have added references to the classic books of Caswell and Morris & Doak.

@@

L222-224: Cite introduction or thorough analyses of these models to direct the reader to more resources. Or add those citations to Table 1 and reference it here for brevity.

>>>Thank you for raising this point, as it spurred us to add clarity about where these models come from. These models are all described by Cohen (1995), who also cites the original sources for each model. Details of each model are also given in Appendix S2. These facts are now mentioned at the point in the text to which the referee refers. Thanks again.

@@

Brief summary: This paper examined the effects of “tail associations” of ecological variables on the extinction risk of metapopulations and stage-structured populations using nonlinear population models. The results show that the effect of tail associations between fecundity and juvenile survival is minor, while that between patches that a metapopulation occupies increases or decreases extinction risk, depending on the domain of different compensatory dynamics. This work is one of the earliest studies that examine how copula structure in ecological variables influence population persistence. The authors investigated this problem using a variety of models, showing the general importance of tail associations of ecological variables on population persistence. Because copula structure in ecological variables should be ubiquitous in nature and ongoing climate change is likely altering correlation structure between variables, the problem investigated in this paper is of interest to the readers of ESA journals.

>>>Thank you for the positive feedback!

General comments
1.      The title seems a bit cryptic to readers who are not a statistician. I would suggest that title be modified to clarify it is about the tails of the distributions of ecological variables.

>>>Thank you for the suggestion. We changed the title as suggested.

@@

2.      A conceptual schematic diagram early in the introduction will greatly help readers understand the main idea quickly and smoothly follow the text. In the current form, it is relatively hard to follow. Figure 1 does help, but a figure showing the idea in a more concrete way using tangible examples would make it much easier for readers to absorb the idea early on in the paper. It also would be helpful to see an example from finance or other fields that are somewhat analogous to ecological situations.

***Shya: I think Fig. 1 is the most simplified pedagogical fig. we've ever had in our all copula-related papers. 
So, as of now I have no idea what extra info does the 2nd reviewer ask for? Copula related paper from finance or other fields are already cited in the paper in L156-157.
***DAN: I do not think we can dodge this - we have to do SOME or even MOST of what the refere says, or he/she will simply say, if the paper is sent back to him/her, "they did not do what I asked for, so REJECT". Of the different suggestions, I think we should make some effort at this one and at 4 below. I will write a justification below for not doing 3. Please think more about such a figure, ASAP. Please think of an *additional* figure, to come right after the curent Fig. 1 The current Fig. 1 explains what tail dependence is. But I think what the referee wants is a figure explaining, conceptually, how tail dependence could cause accentuated extinction risk. I think he/she wants a conceptual figure showing the main idea. I think the figure should be associated with, and cited from, the Intro paras beginning "There are also reasons to hypothesize..." and "Extinction risk is also very commonly estimated..." I agree with the referee that we can produce a good conceptual figure illustrating this main idea. I know you can do this. Please think about it Monday and come to me Tuesday with a hand-drawn sketch of a figure that we can discuss before you then create the figure on Tuesday. Please do not come back to me saying you do not know how to do this - you can do this, I know you can. And time is getting a little tight, and I have other things I have to do on Monday, so this has to be done by you! I am sure you can do a good job.
***DAN: Part of the reason we need to move efficiently on this is they only gave us three weeks, and we may also have to convert to Word, which you know can be labor intensive. Can you please look on the Ecosphere website and find out if they need final submissions to be in Word or if they can take latex? Even if they can take latex, we may want to convery to Word. If there is any evidence that they prefer Word, we will want to convert to Word to avoid them introducing a million errors trying to do latex when they do not really know how.

@@

3.      The amount of analyses for metapopulations and age-structured populations seems quite unbalanced. Why is this so? Obviously, the paper would greatly increase its value by including more analyses for stage-structured populations, but then the paper may become too long to fit in the required length for Ecosphere. Would it be more beneficial to focus only on metapopulations in this paper and expand on the stage-structured population part in a separate paper to more deeply examine Tuljapurkar’s approximation?

>>>This is a first attempt to understand the impact of asymmetric tail associations on extinction risk, and as such we intentionally used a simple approach, using simulations instead of attempting mathematical analyses, and using simple models, in order to assess if the ideas have merit. We believe this could turn into a broad topic, which many researchers should investigate. Our intention was to provide an initial exploration of some potential mechanisms, and in so doing, hopefully to open what could be a practically important topic (e.g., for conservation) to exploration by other researchers. With regard to mathematical analysis of stochastic matrix models, we are journeymen only, not true experts. We believe attempting to carry out an analytic approach ourselves, e.g., using Tuljapurkars approximation, as the referee suggests, would actually slow down the scientfic progress of the field relative to what progress may be possible if stochastic matrix model experts see our curent results and decide to start from the ideas we have posited and then to carry out an appropriate mathematical analysis that would test the generality of our matrix model results (and related models such as integral projection models, which are becoming more and more common and which require even more sophistication to mathematically analyze). We agree with the referee that a systematic, analytic treatment of stochastic matrix models with asymmetric tail associations between vital rates is a worthy goal, but we think the best/fastest way to achieve this is to publish our current result, in hopes of stimulating some appropriate experts to carry out such an analysis. Carving our matrix models results out of the current paper and then undertaking a more systematic explortion ourselves, for possible later publication, does not get these potentially practically important ideas out there as fast. Finally, we do not see such a serious imbalance - to us the two components somewhat, but not dramatically different in the extent of their coverage, not so much for us to want to slow the transmission to the research community of what could be practically important ideas. Certainly our matrix modelling results are enough to raise the possibility that tail associations in vital rates may influence extinction risk, but to suggest that this effect may be minor; and transmitting this idea was our goal. A truly systematic exploration of that idea, using analysis, will require a real expert in the field.

***DAN: Dan will edit and shorten the above.

### 2020-01-19 MainText.Rmd
worked on comments and edits

---
title: "Tail associations and their impact on extinction risk"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Daniel C. Reuman"
date: ""
fontsize: 12pt
geometry: "left=1in,right=1in,top=1in,bottom=1in"
output: 
  pdf_document:
    number_sections: no
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_maintext.sty

@@

<!--stochastic matrix model : age-structured-->
To investigate Q1 from the Introduction,
we considered a simple stochastic age-structured model
with two age classes, 
\begin{equation}
\left({\begin{array}{cc} E(t+1) \\ A(t+1) \end{array}}\right) = 
\left(\begin{array}{cc} 0 & f_a(t)\\ s_e(t) & 0 \end{array}\right)
\left(\begin{array}{c} E(t) \\ A(t) \end{array}\right), \label{eq_age_str}
\end{equation}
\noindent where $E(t)$ and $A(t)$ are the numbers or population densities of eggs 
and adults, respectively, at time $t$. We chose a very simple model
because our goal was to carry out an initial evaluation, only, of 
the realism of a potential mechanism. The formulation (\ref{eq_age_str}) is quite common 
in population-demography modeling [@Caswell2000; @MorrisDoak2002] where vital rates 
are found to be time-dependent and contains stochasticity (due to random 
environmental noise) in population projection matrix. The stochastic vital 
rates $s_e(t)$ and $f_a(t)$ in Eq. \ref{eq_age_str} were selected, independently 
across times, as follows. The egg survival rate $s_e(t)$ was 
drawn from a beta distribution with parameters $\alpha=3$ and $\beta=2$. 
Adult fecundity was $f_a(t)=\tilde{f}_a(t)\exp(-A(t)/\Omega)$, where $\tilde{f}_a(t)$
was drawn from a gamma distribution with shape and scale parameters $k=2$
and $\theta=1$, respectively, 
and $\Omega$ is a density dependence parameter. Taking $\Omega=\infty$ made
$\exp(-A(t)/\Omega)=1$ so that the model did not include density dependence
(it was then a standard stochastic matrix model). Finite, positive values of 
$\Omega$ made the model nonlinear, with density dependence in the adult
fecundity rate.

@@

<!--density-dependence metapopulation model-->
To investigate Q2 from the Introduction, we considered spatialized, stochastic versions
of six density dependent population models, the Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, 
and Austin-Brewer models - as adapted from @cohen1995unexpected.
All models took the general form $\vec{P} (t+1) = D \lambda(t) \vec{P}(t)$,
where $\vec{P}(t)$ is a vector of population densities, 
the $i^{\text{th}}$ component $P_i(t)$ representing
the population density in the $i^{\text{th}}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{\text{th}}$ diagonal entry 
$\lambda_i(t)$ depending on the model used (Table \ref{tab_model_summary}). The matrix $D$
was an $n \times n$ dispersal matrix. For simplicity, 
we assumed the $n$ habitat patches were arranged, evenly spaced, in a line.
We considered both "local" and "global" dispersal among patches in different simulations. 
For local dispersal, a fraction $d$ of 
each population dispersed during each time step, equally distributed to
the two or one nearest neighbors. For global dispersal the same fraction $d$ dispersed
equally to the other $n-1$ patches. When $d$ was $0$, dispersal did not occur and 
rescue effects were not possible.

>>>>

---
title: "Tail associations in ecological variables and their impact on extinction risk"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Daniel C. Reuman"
date: ""
fontsize: 12pt
geometry: "left=1in,right=1in,top=1in,bottom=1in"
output: 
  pdf_document:
    number_sections: no
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_maintext.sty

@@

<!--stochastic matrix model : age-structured-->
To investigate Q1 from the Introduction,
we considered a simple stochastic age-structured model
with two age classes, 
\begin{equation}
\left({\begin{array}{cc} E(t+1) \\ A(t+1) \end{array}}\right) = 
\left(\begin{array}{cc} 0 & f_a(t)\\ s_e(t) & 0 \end{array}\right)
\left(\begin{array}{c} E(t) \\ A(t) \end{array}\right), \label{eq_age_str}
\end{equation}
\noindent where $E(t)$ and $A(t)$ are the numbers or population densities of eggs 
and adults, respectively, at time $t$. We chose a very simple model
because our goal was to carry out an initial evaluation, only, of 
the realism of a potential mechanism. The stochastic vital 
rates $s_e(t)$ and $f_a(t)$ in Eq. \ref{eq_age_str} were selected, independently 
across times, as follows. The egg survival rate $s_e(t)$ was 
drawn from a beta distribution with parameters $\alpha=3$ and $\beta=2$. 
Adult fecundity was $f_a(t)=\tilde{f}_a(t)\exp(-A(t)/\Omega)$, where $\tilde{f}_a(t)$
was drawn from a gamma distribution with shape and scale parameters $k=2$
and $\theta=1$, respectively, 
and $\Omega$ is a density dependence parameter. Taking $\Omega=\infty$ made
$\exp(-A(t)/\Omega)=1$ so that the model did not include density dependence
(it was then a standard stochastic matrix model). Finite, positive values of 
$\Omega$ made the model nonlinear, with density dependence in the adult
fecundity rate. Stochastic matrix models are very commonly used in
evolutionary ecology and quantitative conservation biology, so our model
is an example of a very widely used category of models. See, for instance,
the classic books of @Caswell2000, and @MorrisDoak2002.

@@

<!--density-dependence metapopulation model-->
To investigate Q2 from the Introduction, we considered spatialized, stochastic versions
of six density dependent population models, the Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, 
and Austin-Brewer models. These models were all described by @cohen1995unexpected. 
The models are also listed in Table \ref{tab_model_summary}, and 
are analyzed in detail in Appendix \ref{SM-model_stability}, where 
references for the models are also provided.
All models took the general form $\vec{P} (t+1) = D \lambda(t) \vec{P}(t)$,
where $\vec{P}(t)$ is a vector of population densities, 
the $i^{\text{th}}$ component $P_i(t)$ representing
the population density in the $i^{\text{th}}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{\text{th}}$ diagonal entry 
$\lambda_i(t)$ depending on the model used (Table \ref{tab_model_summary}). The matrix $D$
was an $n \times n$ dispersal matrix. For simplicity, 
we assumed the $n$ habitat patches were arranged, evenly spaced, in a line.
We considered both "local" and "global" dispersal among patches in different simulations. 
For local dispersal, a fraction $d$ of 
each population dispersed during each time step, equally distributed to
the two or one nearest neighbors. For global dispersal the same fraction $d$ dispersed
equally to the other $n-1$ patches. When $d$ was $0$, dispersal did not occur and 
rescue effects were not possible.
